I can see clearly now: reinterpreting statistical significance

### 2019-01-13 doc.tex
few comments by MK

The value of NHST is that asking whether we can reject the null hypothesis is a proxy for asking whether we see clearly \emph{how} our data differs from it. For example, in a t-test, we are nominally asking whether we can see a \emph{difference} between two means, but the scientific question is whether we are confident which of the two means is \emph{larger}; similarly, tests for whether two values are correlated are a proxy for whether we are confident about the \emph{sign} of the correlation coefficient \citep{robinson2001past}. In other cases (e.g., a one-way ANOVA), it may not be simple to describe the difference we see, but NHST is still a reasonable, widely accepted way to evaluate whether an effect has been seen clearly.

@@

\noindent Researchers often write sentences like, ``X et al. showed that there is no significant effect of Y on Z'' with the implication that this effect can now be assumed to be absent (or unimportant). In fact, the sentence is erroneous even before we get to the implication: significance tests provide information about \emph{a data set} -- that is, about a study, not about the study system \citep{HoenigandHeisey2001}. Indeed, a very small effect can lead to $p < 0.05$, when data is abundant (or noise is small); or a very large one can lead to $p > 0.05$ when the sample is small or noisy.

The statement ``X et al. showed that Y has a statistically significant effect on Z'' is similarly misleading. Frequentist statistics effectively assume that the effect is present (or at least, admit that it can't be disproven). The question is whether it is seen in a particular data set. The statement ``X et al. were able to see the effect of Y on Z'' is not only more accurate, but it appropriately implies that something is missing: \emph{What} effect did they see?

@@

\noindent The language of ``statistical clarity'' could help researchers escape various logical traps while interpreting the results of NHST, allowing for the continued use of NHST as a simple, robust method of evaluating whether a data signal is clear (see \cite{Abelson1997} for arguments for NHST). The use of ``significance'' to describe the results of hypothesis tests is deeply, and sometimes subtly, misleading, because it is at odds with other meanings of the word: the \pval is not an accurate gauge of whether a result is large in magnitude, biologically important, or relevant. ``Clarity,'' on the other hand, is an apt term for what NHST actually evaluates. \citet{jones2000sensible} and \citet{robinson2001past} suggest that researchers should report $p > 0.05$ using language such as ``the direction of the differences among the treatments was undetermined''. This is a step in the right direction. Replacing ``significance'' with ``clarity'' takes this idea further, and has the potential to improve statistical communication.

@@

Changing from ``significance'' to ``clarity'' should help researchers improve their statistical practice, but of course it cannot solve all of our problems with statistics. \bmb{repetitive with \P 3 of Conclusion?} The idea of ``statistical clarity'' will work best if it remains linked to the principles of attaching clarity to the study, rather than the system, and of focusing on effects and confidence intervals. If widely adopted, ``statistical clarity'' could eventually come to be seen as an end in itself, the way that ``significance'' is now. We hope not, but in we feel that the unthinking use of ``clarity'' would be (marginally) better than the current unthinking use of ``significance''. If there is a transition it will also be important to communicate clearly when ``clarity'' is being used in a technical sense; we have found that in particular that understanding is improved by explicitly connecting clarity statements to statements about P values.

>>>>

The value of NHST is that \mk{Maybe just begin this sentence with ``Asking''? I still don't really like how this sentence sounds} asking whether we can reject the null hypothesis is a proxy for asking whether we see clearly \emph{how} our data differs from it. For example, in a t-test, we are nominally asking whether we can see a \emph{difference} between two means, but the scientific question is whether we are confident which of the two means is \emph{larger}; similarly, tests for whether two values are correlated are a proxy for whether we are confident about the \emph{sign} of the correlation coefficient \citep{robinson2001past}. In other cases (e.g., a one-way ANOVA), it may not be simple to describe the difference we see, but NHST is still a reasonable, widely accepted way to evaluate whether an effect has been seen clearly.

@@

\noindent Researchers often write sentences like, ``X et al. showed that there is no significant effect of Y on Z'' with the implication that this effect can now be assumed to be absent (or unimportant). In fact, the sentence is erroneous even before we get to the implication: significance tests provide information about \emph{a data set} -- that is, about a study, not about the study system \citep{HoenigandHeisey2001}. Indeed, a very small effect can lead to $p < 0.05$, \mk{remove Indeed and both commas in this sentence (semi-colon becomes a comma)?} when data is abundant (or noise is small); or a very large one can lead to $p > 0.05$ when the sample is small or noisy.

The statement ``X et al. showed that Y has a statistically significant effect on Z'' is similarly misleading. Frequentist statistics effectively assume that the effect is present (or at least, \mk{no comma needed here?} admit that it can't be disproven). The question is whether it is seen in a particular data set. The statement ``X et al. were able to see the effect of Y on Z'' is not only more accurate, but it appropriately implies that something is missing: \emph{What} effect did they see?

@@

\noindent The language of ``statistical clarity'' could help researchers escape various logical traps while interpreting the results of NHST, allowing for the continued use of NHST as a simple, robust method of evaluating whether a data signal is clear (see \cite{Abelson1997} for arguments for NHST) \mk{Is there a better place to cite this (maybe in the last sentence of the intro)?}. The use of ``significance'' to describe the results of hypothesis tests is deeply, and sometimes subtly, misleading, because it is at odds with other meanings of the word: the \pval is not an accurate gauge of whether a result is large in magnitude, biologically important, or relevant. ``Clarity,'' on the other hand, is an apt term for what NHST actually evaluates. \citet{jones2000sensible} and \citet{robinson2001past} suggest that researchers should report $p > 0.05$ using language such as ``the direction of the differences among the treatments was undetermined''. This is a step in the right direction. Replacing ``significance'' with ``clarity'' takes this idea further, and has the potential to improve statistical communication.

@@

Changing from ``significance'' to ``clarity'' should help researchers improve their statistical practice, but of course it cannot solve all of our problems with statistics. \bmb{repetitive with \P 3 of Conclusion?} \mk{Maybe not needed here at all considering that this second sentence doesn't even directly follow. I vote to just remove this first sentence} The idea of ``statistical clarity'' will work best if it remains linked to the principles of attaching clarity to the study, rather than the system, and of focusing on effects and confidence intervals. If widely adopted, ``statistical clarity'' could eventually come to be seen as an end in itself, the way that ``significance'' is now. We hope not, but in we feel that the unthinking use of ``clarity'' would be (marginally) better than the current unthinking use of ``significance''. If there is a transition it will also be important to communicate clearly when ``clarity'' is being used in a technical sense; we have found that in particular that understanding is improved by explicitly connecting clarity statements to statements about P values.
